The Weakly Nonlinear Magnetorotational Instability

### 2016-08-31 paper/widegap_mri_paper.tex
added some discussion

Our work should be placed in the broader context of emergent pattern formation in physical systems.

@@

>>>>

Our work should be placed in the broader context of emergent pattern formation in physical systems. The real Ginzburg-Landau equation derived here governs the slow-parameter evolution of the standard MRI close to threshold. The GLE arises in a number of other physical systems, and in each case it is a consequence not of the particular physics at hand, but of the underlying symmetries in the problem. Here we make a brief comparison to two other systems that give rise to a GLE. The first and perhaps most famous is Rayleigh-B\'enard convection, in which a fluid between two plates is heated from below. If we take the plane of the fluid to be infinite in the horizontal plane, the system is initially translationally symmetric. %When the system eventually becomes unstable to convective heat transfer it undergoes a symmetry breaking, allowing the 
At the onset of convection the system undergoes a symmetry breaking, forming rolls, or convection cells, which break the horizontal translational invariance. Analogously, the standard MRI system considered here is initially vertically translationally symmetric, because we idealize the Taylor-Couette device as an infinitely long cylinder. The MRI breaks this symmetry, forming cells along the vertical length of the domain. Just as Rayleigh-B\'enard cells transport heat vertically, the MRI cells transport angular momentum horizontally. The symmetry breaking of each of these systems is described near onset by the real GLE.

A real GLE has also been found to describe the formation of zonal flows out of magnetized turbulence in a model system (). Zonal flows are axisymmetric structures, large-scale and long-lived, which form spontaneously out of turbulence. They have recently been observed in some numerical studies of the MRI, and have generated considerable interest for their possible role in planet formation in protoplanetary disks (). The present work is of course an idealized geometry, and we make no attempt to model a realistic protoplanetary disk environment. However, it is worth noting that the GLE we derive implies that axisymmetric, large-scale, long-lived structures are a generic feature of the MRI in the weakly nonlinear regime. This work provides a mathematical description of the MRI as a pattern-forming process, but much remains to be understood, particularly involving the relevance of these insights to realistic astrophysical disks. Our model is most directly relevant to Taylor-Couette flows, and we emphasize that laboratory MRI experiments stand poised to observe the MRI-driven pattern formation predicted here.

@@

We solve the $\mathcal{O}(\epsilon)$ (linear) system, followed by the $\mathcal{O}(\epsilon^2)$ system in Equation \ref{eq:ordere2}. At second order in $\epsilon$, nonlinear terms arise which are formed by the interaction of first-order MRI modes with themselves and their complex conjugates. This mode interaction means that the second-order nonlinear term is

\beq
\mathbf{N}_2 = |\alpha|^2 \mathbf{N}_{20} + \alpha^2 \mathbf{N}_{22} e^{2 i k_c z},
\eeq

where terms are grouped by z-dependence. See Appendix \ref{app:nonlinear_terms} for the full form of the nonlinear terms. Equation \ref{eq:ordere2} must therefore be solved as three separate systems of equations, one for each possible $z$ resonance:
